A Bidirected String Graph Model for Genome Assembly

### 2012-12-13 presentation/paper.tex
Figures throughout
paper: finalize

\usepackage[nofiglist,notablist]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}

@@

Based on these biological facts, a genome is formally defined, for the purposes
of this paper, as a set of possibly circular, dual-stranded strings of the
symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
meaning indicated above (the two strands run in opposite directions, and the
pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold).

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the whole genome must be broken into much smaller fragments,
which are then sequenced.\footnote{This specifically refers to the
\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
restrict sequences to certain regions of the genome in order to make sequencing
the genome easier (currently, at the disadvantage of increased cost and much more
non-computational work).} Following this, the original genome must be
computationally reconstructed by a computer program called a \KeyTerm{genome
assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
sequencing} therefore refers to the entire process of sequencing a genome,
including both lab work and computational work, while \KeyTerm{genome assembly}
only refers to the computational work.\footnote{It usually also takes some
technology-specific computational work to actually prepare the sequences for
assembly; for example, a series of digital photographs of clumps of
fluorescently labeled molecules might need to be interpreted as reads using
image-processing algorithms.  However, these programs and/or algorithms
generally have already been developed by the manufacturer of the sequencing
machine and are assumed to work fairly well, so ``genome assembly'' does not cover these
algorithms.  Still, assemblers may be tuned for specific sequencing
technologies.}

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from condition (4) is
{\em not} a reverse-complement overlap, as it can equivalently be stated as the
last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$ bp of $f$.
\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
may arise from this definition.

@@

Overlaps are important because two reads that overlap may have come from
adjacent positions on the genome.  Although due to repeats, we cannot be certain
that two overlapping reads actually came from adjacent positions, we {\em can}
say for certain that two reads that do not overlap did not come from adjacent
positions.  Regardless of this, \Figure{fig:overlap_examples} shows how two
reads that overlap potentially could be combined to form a longer sequence.

@@

Now that the idea of an exact overlap has been formalized, what we would like to
do is compute all pairwise exact overlaps of some minimum length $L$ among the
	reads that were provided as input to the assembler.  The chosen value of $L$
	must be balanced so that we can find enough overlaps to adequately cover the
	genome (small $L$ better) but still exclude most spurious overlaps (large
	$L$ better).

The naive way to compute overlaps is to consider every pair of reads in the
input, including only distinct pairs but also including the special cases of
each read paired with itself, and search for each of the types of overlaps
between the two reads in each pair.  If there are $N$ reads in the input, this
algorithm will need to consider $\frac{N^2}{2} + N$ pairs of reads. The time
spent on each read pair will depend several factors, including the minimum
overlap length $L$ and the read lengths.  But even if we somehow could spend
constant time on each read pair, this naive algorithm would be prohibitively
slow because the running time rises quadraticly with respect to the number of
reads, and there can easily be millions of reads in the input.

A number of faster algorithms for computing overlaps have been designed.  One of
the easier ones to understand is the \KeyTerm{seed-and-extend} algorithm.  It
works by indexing the reads by short subsequences of a certain length (less than
or equal to than the minimum overlap length $L$) and then only comparing reads
that actually share a subsequence.  Each of these short subsequences is called a
\KeyTerm{seed}.  Two reads may share a seed in either a forward or
reverse-complement manner.  Given two reads that share a seed, the seed can be
extended outwards in both directions to find a maximal matched region (that may
be either forward or reverse-complement) between the two reads.  Then, this
maximal matched region is compared against the definition of an overlap.  In
particular, it must be at least as long as the minimum overlap length, it must
have least 2 extreme points over the two reads, and it cannot be the prefix or
suffix of both reads if the match does not extend to the opposite end of at
least one of the reads.  If all of these conditions are met, this maximal matched
region is translated into an overlap and added to the set of overlaps if not
already present.

@@

For this project, the seed-and-extend algorithm for computing overlaps was
implemented as the program \ProgramName{compute-overlaps}.  It takes as input a
file that contains a list of reads, as well as a parameter that specifies the
minimum overlap length $L$.  It produces as output (via a data structure
serialized to a file) all exact pairwise overlaps of minimum length $L$ among
reads in the input file, including reverse-complement overlaps and
self-overlaps.  The running time of the algorithm was discussed above.  The
memory usage is linear with respect to the total length of the reads and the
number of overlaps, which likely could be improved.

@@

A \KeyTerm{walk} in a bidirected graph $G$ is a continuous sequence of edges in
$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
through a head outwards (unless it is the end of the path), and vice versa.  An
equivalent formulation is that if we consider a possible walk in a bidirected
graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
not, we must consider every vertex in the possible walk except for the very
first and last vertices.  For each such vertex, we need to verify that each time
that vertex is passed through as part of a walk, the arrow heads adjacent to
that vertex that are used in the relevant part of the walk are pointed in the
same direction {\em relative to the walk}. If there are no such pairs of heads
that point towards each other or point away from each other, the walk is valid.
The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.

@@

After transitive reduction, a non-repetitive region of the genome will typically
be represented in the string graph by a chain of vertices that is part of a
valid walk and consists entirely of vertices with in-degree 1 and out-degree 1.
This is referred to as an \KeyTerm{unbranched path}, and it contains one or more
\KeyTerm{inner vertices}.  An unbranched path can be collapsed by removing the
inner vertices and replacing them by a single edge that covers the entire path.
This replacement edge is labeled with the concatenation of the sequences that
the edges are labeled with.  In the bidirected case, this concatenation must be
done for both directions in which the unbranched path might have been traversed.
\Figure{fig:collapse_unbranched_paths} shows an example of how a short
unbranched path can be collapsed.

@@

		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node (5) at (3.25, 0) {5};
			\node (6) at (4.75, 1) {6};
			\node (7) at (4.75, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}
	\end{center}
	\caption{Vertex 4 an \KeyTerm{inner vertex} that is part of an unbranched
	path that can be collapsed.  Edge labels are not shown.}
	\label{fig:collapse_unbranched_paths}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and unbranched path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

Real genomes are not random, and they contain far more repeats than would be
expected purely by chance.\footnote{Biologically, repeats can occur for a number
of reasons, such as gene duplications.} \Figure{fig:E_coli} shows an example of
the simplified bidirected string graph for a real genome.  It is apparent that
even when using perfect, fairly long reads that cover the entire genome, the
repeats in the genome cause branches to appear in the graph when the repeats are
too long to be resolved by the reads at this stage in the assembly.

@@

This section walks through the construction and simplification of the bidirected
string graph on a very small example and provides a more detailed look at the
actual data.  The genome is 200 bp long and is given by the following file:

@@

Note that in this file format, a continuous sequence can be split among multiple
lines; a new sequence only begins after a line beginning with {\tt >} is
reached.

@@

	\fbox{
	\begin{tikzpicture}[>=\ReadArrowType,x=0.56mm,y=3.5mm]
		\draw[<-, style=thick] (0, 1) node[anchor=east] {G.} -- (200, 1);
		\draw[->, style=thick] (0, 0) node[anchor=east] {G.} -- (200, 0);

@@

These overlaps are then used by \ProgramName{build-bidirected-string-graph} to
build the assembly graph.  Note that there are no contained reads in this
example, so there are as many vertices as original reads.  Also, in the below
graph drawing (as well as following graph drawings), each edge is labeled with
the length of that edge's sequence rather than the sequence itself.
Furthermore, although each edge really contains a separate sequence for each
direction, for this particular example all the reads are the same length and
only exact overlaps are considered, which makes it so the two sequences of each
edge are always the same length.

@@

The result is the fully simplified bidirected string graph.  In this particular
example, the final graph is trivial and can be used to assemble the read-sampled
part of the original genome by walking $2 \to 1$.  Note that there is an implied
50bp before we start at the end of the read now labeled as 2, so the actual
assembled sequence is $113 + 50 = 163$ bp long.  The assembler should then
generate the assembly:

@@

A contained read may easily be mapped to the vertex corresponding to a read that
contains it, but this is not sufficient because the A-statistic is calculated
over the edges, not the vertices.  Instead, we need to map each end of each
contained read to the edge, or in general the set of edges, in which the end
might be located if it were included as its own vertex.

The algorithm for doing so is outlined in \cite{Myers2005} and was implemented
by the program \ProgramName{map-contained-reads}, which runs before
\ProgramName{collapse-unbranched-paths}.  Briefly, for each contained read $r$
the algorithm identifies the uncontained read $R$ that contains $r$ and has the
shortest distance from the end of $r$ to the end of $R$ (the shortest
\KeyTerm{overhang}).  This is done for each of the two directions.  The edges in
the graph starting at the vertex corresponding to that end of $R$ are then
walked backwards for the length of the overhang to find the set of edges $S$ to
which the relevant end of $r$ may map, and each edge in $S$ is marked as
containing $\frac{1}{|S|}$ extra inner vertices.  This mapping, therefore, makes
it possible to account for contained reads when calculating the A-statistic.

@@

Formally, a minimum-cost circulation problem on a directed graph takes as input
for each edge $e$ an lower bound $l(e) \ge 0$ and an upper bound $u(e) \ge 0$ on
	flow and a cost per unit flow $v(e)$, and for each vertex $v$ a supply
	(positive) or demand (negative) $b(v)$ for flow.  It produces as output a
	flow $x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized
	subject to $x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) =
	\sum_{v \to w} x(v \to w)$ \cite{Myers2005}.  In the minimum-cost
	circulation problem, as opposed to other network flow problems, no vertices are
	treated specially as sources or sinks; instead, flow must be balanced at
	every vertex $v$, but only after taking into account the supply or demand
	$b(v)$ at that vertex.

The A-statistic calculation on each edge $e$, as described in
\Subsection{subsec:a_statistics}, allows us to assign a lower bound $l(e)$ and
an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
may be $\infty$.  We also have $b(v) = 0$ for all $v$, because the flow at every
vertex must be balanced; this corresponds to the desired walk entering each
vertex the same number of times that it is exited, making the graph as a whole
be balanced.  Furthermore, we have $v(e) = 1$ because no edges are considered to
be more costly to use than any other in this formulation.\footnote{Note that the
goal of this assembly algorithm is not to minimize the assembled genome length,
but rather to maximize the similarity of our assembled genome to the true
genome.  Therefore, there is no need to assign greater costs to edges that
represent longer DNA sequences (at least in this formulation of the problem),
because that would make it more costly to repeat a longer edge even if there is
strong evidence that edge corresponds to repeated sequence.}

@@

There are still two serious problems with this circulation formulation that need
to be addressed.  First, we actually have a bidirected graph, not a directed
graph, so we need to calculate a \KeyTerm{bidirected flow}.\footnote{The fact
that this circulation problem actually needs to be formulated and solved on a
bidirected graph is addressed in \cite{Medvedev2009}, but not in
\cite{Myers2005}, even though the latter introduced the overall fragment string
graph algorithm.}
A bidirected flow is not the same as a directed flow because of the
presence of \BidirectedEdgeInnie and \BidirectedEdgeOutie edges, where the
magnitude of the flow on that edge is counted as going into both vertices (in
the case of a \BidirectedEdgeOutie edge) or out of both vertices (in the case of
an \BidirectedEdgeInnie edge).  This does not imply that the edge itself is a
source or a sink, but rather than the flow is defined in a peculiar way that
makes sense on bidirected graphs.

@@

The second problem with our circulation formulation is that the bidirected
string graph is not guaranteed to be connected, let alone strongly connected.
For example, there might be vertices that have in-degree 0 but have one or more
adjacent out-edges with nonzero lower bound on flow.  Such vertices may
represent the last read in a genomic region before a coverage gap or the end of
the sequence.  Our goal is to have such vertices be the endpoints of logical
walks through the bidirected graph.  To take this into account in the
formulation of the bidirected flow problem, we can introduce a special vertex
$s$ and unlabeled special edges between $s$ and every other vertex.  Each
special edge $e$ can have any amount of flow ($l(e) = 0,\ u(e) = \infty$), but
at great cost ($v(e) = X$, where $X$ is a very large number).  The addition of
these edges will make the graph strongly connected, and therefore it will be
possible to solve the minimum-cost circulation problem to find a flow that
corresponds to balancing the in-degree and out-degree of every vertex by
assigning a certain traversal count $t(e)$ to each edge $e$.

@@

Here, a \KeyTerm{generalized Eulerian cycle} means a cycle in a bidirected graph
$G$ that traverses each edge $e$ exactly $t(e)$ times, where $t(e)$ is the flow
on edge $e$ as calculated using the method in \Subsection{subsec:circulation}.
A generalized Eulerian cycle is guaranteed to exist in the bidirected graph at
this point because the graph was made strongly connected before solving the
minimum-cost circulation problem.  (A more thorough explanation of this
observation can be found in \cite{Medvedev2009}.)  The cycle corresponds to a
consistent assembly of the genome, and it can be decomposed into a set of
logical walks by placing a logical breakpoint every time the special vertex $s$
is passed through.  Each of these logical walks represents the assembly of a
separate contig.

@@

\begin{itemize}
\item Finish the minimum-cost circulation and generalized Eulerian path parts of
the algorithm so that the implementation is ``complete'' and can produce a
``final'' assembly.
\item Improve the performance of \ProgramName{compute-overlaps}.
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are inexact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this
algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
Burrows-Wheeler transform to efficiently construct the transitively-reduced
bidirected string graph directly from the reads \cite{Simpson2010}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence, possibly using ideas
from \cite{Medvedev2009}.
\end{itemize}

>>>>

%\usepackage[nofiglist,notablist]{endfloat}
%\renewcommand{\efloatseparator}{\mbox{}}

@@

Based on these biological facts, a genome is formally defined, for the purposes
of this paper, as a set of possibly circular, double-stranded strings of the
symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where double-stranded has
the meaning indicated above (the two strands run in opposite directions, and the
pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold).

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the whole genome must be broken into much smaller fragments,
which are then sequenced.\footnote{This specifically refers to the
\KeyTerm{whole-genome shotgun sequencing} approach to genome sequencing rather
than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
restrict sequences to certain regions of the genome in order to make sequencing
the genome easier (currently, at the disadvantage of increased cost and much more
non-computational work).} Following this, the original genome must be
computationally reconstructed by a computer program called a \KeyTerm{genome
assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
sequencing} therefore refers to the entire process of sequencing a genome,
including both lab work and computational work, while \KeyTerm{genome assembly}
only refers to the computational work.\footnote{It usually also takes some
technology-specific computational work to actually prepare the sequences for
assembly; for example, a series of digital photographs of clumps of
fluorescently labeled molecules might need to be interpreted as reads using
image-processing algorithms.  However, these programs and/or algorithms
generally have already been developed by the manufacturer of the sequencing
machine and are assumed to work fairly well, so ``genome assembly'' does not cover these
algorithms.  Still, assemblers may be tuned for specific sequencing
technologies.}

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from the last
condition is {\em not} a reverse-complement overlap, as it can equivalently be
stated as the last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$
bp of $f$.  \Figure{fig:overlap_types} shows the three fundamental types of
overlaps that may arise from this definition.

@@

Overlaps are important because two reads that overlap may have come from
adjacent positions on the genome.  Although due to repeats, we cannot be certain
that two overlapping reads actually came from adjacent positions, we {\em can}
be certain that two reads that do not overlap did not come from adjacent
positions.  Regardless of this, \Figure{fig:overlap_examples} shows how two
reads that overlap potentially could be combined to form a longer sequence.

@@

Now that the idea of an exact overlap has been formalized, our goal is to
compute all pairwise exact overlaps of some minimum length $L$ among the reads
that were provided as input to the assembler.  The chosen value of $L$ must be
balanced so that we can find enough overlaps to adequately cover the genome
(small $L$ better) but still exclude most spurious overlaps (large $L$ better).

The naive way to compute overlaps is to consider every pair of reads in the
input, including only distinct pairs but also including each read paired with
itself, and search for each of the types of overlaps between the two reads in
each pair.  If there are $N$ reads in the input, this algorithm will need to
consider $\frac{N^2}{2} + N$ pairs of reads. The time spent on each read pair
will depend several factors, including the minimum overlap length $L$ and the
read lengths.  But even if we somehow could spend constant time on each read
pair, this naive algorithm would be prohibitively slow because the running time
rises quadraticly with respect to the number of reads, and there can easily be
millions of reads in the input.

A number of faster algorithms for computing overlaps have been designed.  One of
the easier ones to understand is the \KeyTerm{seed-and-extend} algorithm.  It
works by indexing the reads by short subsequences of a certain length (less than
or equal to than the minimum overlap length $L$) and then only comparing reads
that actually share a subsequence.  Each of these short subsequences is called a
\KeyTerm{seed}.  Two reads may share a seed in either a forward or
reverse-complement manner.  Given two reads that share a seed, the seed can be
extended outwards in both directions to find a maximal matched region (that may
be either forward or reverse-complement) between the two reads.  Then, this
maximal matched region is compared against the definition of an overlap.  In
particular, it must have length greater than or equal to $L$, it must have least
2 extreme points over the two reads, and it cannot be the prefix or suffix of
both reads if the match does not extend to the opposite end of at least one of
the reads.  If all of these conditions are met, this maximal matched region is
translated into an overlap and added to the set of overlaps if not already
present.

@@

For this project, the seed-and-extend algorithm for computing overlaps was
implemented as the program \ProgramName{compute-overlaps}.  It takes as input a
file that contains a list of reads and a parameter that specifies the minimum
overlap length $L$.  It produces as output, via a data structure serialized to a
file, all exact pairwise overlaps of minimum length $L$ among the input reads,
including reverse-complement overlaps and self-overlaps.  The running time of
the algorithm was discussed above.  The memory usage is linear with respect to
the total length of the reads and the number of overlaps, which likely could be
improved.

@@

A \KeyTerm{walk} in a bidirected graph $G$ is a continuous sequence of edges in
$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
through a head outwards (unless it is the end of the path), and vice versa.  An
equivalent formulation is that if we consider a possible walk in a bidirected
graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
not, we must consider every vertex in the possible walk except for the first and
last vertices.  For each such vertex, we need to verify that each time that
vertex is passed through as part of a walk, the arrow heads adjacent to that
vertex that are used in the relevant part of the walk are pointed in the same
direction {\em relative to the walk}. If there are no such pairs of heads that
point towards each other or point away from each other, the walk is valid.  The
red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads that
need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.

@@

After transitive reduction, a non-repetitive region of the genome typically will
be represented in the string graph by a chain of vertices that is part of a
valid walk and consists entirely of vertices with in-degree 1 and out-degree 1.
This is referred to as an \KeyTerm{unbranched path}, and it contains one or more
\KeyTerm{inner vertices}.  An unbranched path can be collapsed by removing the
inner vertices and replacing them by a single edge that covers the entire path.
This replacement edge is labeled with the concatenation of the sequences that
the original edges were labeled with.  In the bidirected case, this
concatenation must be done for both directions in which the unbranched path
might have been traversed.  \Figure{fig:collapse_unbranched_paths} shows an
example of how a short unbranched path can be collapsed.

@@

		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node (5) at (3.25, 0) {5};
			\node (6) at (4.75, 1) {6};
			\node (7) at (4.75, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}
	\end{center}
	\caption{Vertex 4 is an inner vertex that is part of an unbranched path that
	can be collapsed.  Edge labels are not shown.}
	\label{fig:collapse_unbranched_paths}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {1};
				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {2};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and unbranched path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

Real genomes are not random, and they contain far more repeats than would be
expected purely by chance.\footnote{Biologically, repeats can occur for a number
of reasons, such as gene duplications.} \Figure{fig:E_coli} shows an example of
the simplified bidirected string graph for reads sampled from a real genome.  It
is apparent that even when using perfect, fairly long reads that cover the
entire genome, the repeats in the genome cause branches to appear in the graph
when the repeats are too long to be resolved by the reads at this stage in the
assembly algorithm.

@@

This section walks through the construction and simplification of the bidirected
string graph on a very small example and provides a more detailed look at the
actual data.  The genome is 200 bp long and is given by the following plain-text
file:

@@

Note that in this file format, a continuous sequence can be split among multiple
lines, and a new sequence only begins after a line beginning with {\tt >} is
reached.

@@

	\fbox{
	\begin{tikzpicture}[>=\ReadArrowType,x=0.56mm,y=3.5mm]
		\draw[<-, style=thick] (0, 1) node[anchor=east] {$G'$.} -- (200, 1);
		\draw[->, style=thick] (0, 0) node[anchor=east] {$G$.} -- (200, 0);

@@

These overlaps are then used by \ProgramName{build-bidirected-string-graph} to
build the assembly graph.  Note that there are no contained reads in this
example, so there are as many vertices as original reads.  Also, in the
following graph drawings, each edge is labeled with the length of that edge's
sequence rather than the sequence itself.  Furthermore, although each edge
really contains a separate sequence for each direction, for this particular
example all the reads are the same length and only exact overlaps are
considered, which makes it so the two sequences of each edge are always the same
length.

@@

The result is the fully simplified bidirected string graph.  In this particular
example, the final graph is trivial and can be used to assemble the read-sampled
part of the original genome by walking $2 \to 1$.  Note that there is an implied
50bp before we start at the end of the read now labeled as 2 (which was
originally read 5), so the actual assembled sequence is $113 + 50 = 163$ bp
long.  The assembler should then generate the assembly:

@@

A contained read may easily be mapped to the vertex corresponding to a read that
contains it, but this is not sufficient because the A-statistic is calculated
over the edges, not the vertices.  Instead, we need to map each end of each
contained read to the edge, or in general the set of edges, in which the end
might be located if it were included as its own vertex.  Note that this mapping
makes more sense if we consider it in terms of the directed analogue of the
bidirected graph introduced in \Subsection{subsec:directed}.

The algorithm for doing this mapping is outlined in \cite{Myers2005} and was
implemented by the program \ProgramName{map-contained-reads}, which runs before
\ProgramName{collapse-unbranched-paths}.  Briefly, for each contained read $r$
the algorithm identifies the uncontained read $R$ that contains $r$ and has the
shortest distance from the end of $r$ to the end of $R$ (the shortest
\KeyTerm{overhang}).  This is done for each of the two directions.  The edges in
the graph starting at the vertex corresponding to that end of $R$ are then
walked backwards for the length of the overhang to find the set of edges $S$ to
which the relevant end of $r$ may map, and each edge in $S$ is marked as
containing $\frac{1}{|S|}$ extra inner vertices.  This mapping, therefore, makes
it possible to account for contained reads when calculating the A-statistic.

@@

Formally, a minimum-cost circulation problem on a directed graph takes as input
for each edge $e$ an lower bound $l(e) \ge 0$ and an upper bound $u(e) \ge 0$ on
flow and a cost per unit flow $v(e)$, and for each vertex $v$ a supply
(positive) or demand (negative) $b(v)$ for flow.  It produces as output a flow
$x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized subject to
$x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w}
x(v \to w)$ \cite{Myers2005}.  In the minimum-cost circulation problem, as
opposed to other network flow problems, no vertices are treated specially as
sources or sinks; instead, ingoing and outgoing flow must be balanced at every
vertex $v$, but only after taking into account the supply or demand $b(v)$ at
that vertex.

The A-statistic calculation on each edge $e$, as described in
\Subsection{subsec:a_statistics}, allows us to assign a lower bound $l(e)$ and
an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
may be $\infty$.  We also have $b(v) = 0$ for all $v$, because the flow at every
vertex must be balanced; this corresponds to the desired walk entering each
vertex the same number of times that it is exited, making the graph as a whole
be balanced.  Furthermore, we have $v(e) = 1$ because no edges are considered
more costly to use than any other in this formulation.\footnote{Note that the
goal of this assembly algorithm is not to minimize the assembled genome length,
but rather to maximize the similarity of our assembled genome to the true
genome.  Therefore, there is no need to assign greater costs to edges that
represent longer DNA sequences (at least in this formulation of the problem),
because that would make it more costly to repeat a longer edge even if there is
strong evidence that edge corresponds to repeated sequence.}

@@

There are still two serious problems with this circulation formulation that need
to be addressed.  First, we actually have a bidirected graph, not a directed
graph, so we need to calculate a \KeyTerm{bidirected flow}.\footnote{The fact
that this circulation problem actually needs to be formulated and solved on a
bidirected graph is addressed in \cite{Medvedev2009}, but not in
\cite{Myers2005}, even though the latter introduced the overall fragment string
graph algorithm.}
A bidirected flow is not the same as a directed flow because of the
presence of \BidirectedEdgeInnie and \BidirectedEdgeOutie edges, where the
magnitude of the flow on that edge is incoming to both both vertices (in
the case of a \BidirectedEdgeOutie edge) or outgoing from both vertices (in the case of
an \BidirectedEdgeInnie edge).  This does not imply that the edge itself is a
source or a sink, but rather than the flow is defined in a peculiar way that
makes sense on bidirected graphs.

@@

The second problem with our circulation formulation is that the bidirected
string graph is not guaranteed to be connected, let alone strongly connected.
For example, there might be vertices that have in-degree 0 but have one or more
adjacent out-edges with nonzero lower bound on flow.  Such vertices may
represent the last read in a genomic region before a coverage gap or the end of
the sequence.  Our goal is to have such vertices be the endpoints of logical
walks through the bidirected graph.  To take this into account in the
formulation of the bidirected flow problem, we can introduce a special vertex
$s$ and unlabeled special edges between $s$ and every other vertex.  Each
special edge $e$ can have any amount of flow ($l(e) = 0,\ u(e) = \infty$), but
at great cost ($v(e) = X$, where $X$ is a very large number).  The addition of
these edges will make the graph strongly connected and possible to solve the
minimum-cost circulation problem to find a flow that corresponds to balancing
the in-degree and out-degree of every vertex by assigning a certain traversal
count $t(e)$ to each edge $e$.

@@

Here, a \KeyTerm{generalized Eulerian cycle} means a cycle in a bidirected graph
$G$ that traverses each edge $e$ exactly $t(e)$ times, where $t(e)$ is the flow
on edge $e$ as calculated using the method in \Subsection{subsec:circulation}.
A generalized Eulerian cycle is guaranteed to exist in the bidirected graph at
this point because the graph was made strongly connected before solving the
minimum-cost circulation problem.  (A more thorough explanation of this
observation can be found in \cite{Medvedev2009}.)  The cycle corresponds to an
assembly of the genome that is consistent with the original reads, and it can be
decomposed into a set of logical walks by placing a logical breakpoint every
time the special vertex $s$ is passed through.  Each of these logical walks
represents the assembly of a separate contig.

@@

\begin{itemize}
\item Finish the minimum-cost circulation and generalized Eulerian path parts of
the algorithm so that the implementation is ``complete'' and can produce a
``final'' assembly.
\item Improve the performance of \ProgramName{compute-overlaps}.
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are inexact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this
algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
Burrows-Wheeler transform to efficiently construct the transitively-reduced
bidirected string graph directly from the original reads \cite{Simpson2010}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence, possibly using ideas
from \cite{Medvedev2009}.
\end{itemize}
